---
title: 'Econmetrics II: Exercise 1'
output:
  pdf_document:
      keep_tex: true
      latex_engine: xelatex
  html_document:
    df_print: kable
fontsize: 12pt
colorlinks: true
header-includes:
  - \usepackage{dcolumn}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  message = FALSE,
  comment = "#>"
)
```


```{r, load-packages}
library(dplyr)
library(stargazer)
```

# Econometrics II  


## Exercise 1

(a) **\color{red}{Load the data into Stata and familiarize yourself with the structure and the variables using different descriptive statistics.}**

<!-- Let's glimpse at the structure of the dataset.  -->

```{r, read-in data}

# url <- "https://www.stata.com/data/jwooldridge/eacsap/nls80.dta"
# download.file(url, "nls80.dta", quiet = TRUE)

nls_80 <- haven::read_dta("nls80.dta") # read in `.dta` data

# get the labels of the variables, `Label` in Stata.
var_labels <- sapply(nls_80, function(x) attr(x, "label"))


# tibble::glimpse(nls_80) # or head(nls_80, 5) to view the first 5 rows
```

This dataset has `r ncol(nls_80)` columns and `r nrow(nls_80)` rows. We can look at the descriptions of variables in the following table. 


```{r, description of variables, results="asis"}
nms <- colnames(nls_80)

var_labels <- tibble::tibble(
  "\\textbf{Variable name}" = nms,
  "\\textbf{Variable label}" = var_labels,
  row.names = NULL
)
knitr::kable(var_labels,
  caption = "Description of variables", label = "tab:vd"
)
```


```{r, results='asis'}

stargazer(as.data.frame(nls_80, stringsAsFactors = FALSE),
  title = "Descriptive statistics",
  header = FALSE, no.space = TRUE, align = TRUE,
  label = "tab:ds", summary.logical = TRUE
)

# make married, black, south, and urban explicit dummy/factor
nls_80 <- mutate_at(nls_80, vars(married:urban), as.factor)
```



(b) **\color{red}{Have a closer look at the wage variable (monthly earnings in \$). How does the mean wage compare to the median? Plot the wage distribution. What is the expected wage of a working man if he is married?}**

Mean of **wage** = \$`r mean(nls_80$wage)` and median \$`r median(nls_80$wage)`. 

```{r, distribution of wage, fig.cap="Distribution of monthly earnings"}
hist(nls_80$wage, xlab = "wage (monthly earnings)", main = NULL)
```


```{r, marital-status}
nls_80_married <- subset(nls_80, married == 1)

nls_80_unmarried <- subset(nls_80, married == 0)
```

The **expected wage** of a **married man** is \$`r mean(nls_80_married$wage)`. The difference between the average wage of married and unmarried person is \$`r mean(nls_80_married$wage)- mean(nls_80_unmarried$wage)`. Unmarried person earns way less. 


(c) **\color{red}{Consider the following equation for men under the assumption that all explanatory variables are exogenous}**


\begin{equation}
wage = \beta_0 + \beta_1married + \beta_2high\_iq + \beta_3medium\_edu + \beta_4high\_educ + z\gamma + u\\
\label{mod.1}
\end{equation}



$E(u|married, iq, educ, z)=0$. 


**How do you interpret $\beta_1$?**  

- A married person would get $\beta_1$ more dollars of wage over unmarried one, **on average**, **other factors held intact**. 

**How would you interpret the constant term?**  

- Usually, the intercept tells us what the value of the dependent variable (here wage) would be when all independent variables are zero. However, in this typical regression setup, the intercept captures the average wage of unmarried person, while all other factors are set to zero.  

(d) \color{red}{Run the regression and interpret the regression coefficients. Include all education dummies to the regression model. Why does Stata drop one of the dummies? How does the interpretation of the constant change if you change the included education dummies?}

```{r, create-dummies, include=FALSE}

# high_iq = 1 if iq is > the median value of iq, 0 otherwise. Then, we can give it optional labels.
nls_80 <- mutate(nls_80,
  high_iq = ifelse(iq > median(iq), 1, 0),
  high_iq = factor(high_iq, c(0, 1), c("low", "high"))
)

# education dummies
# medium vs high; let's cut it at 13 years of education---high school completed?
nls_80 <- mutate(nls_80,
  educ_dummy = case_when(educ < 13 ~ 0, 
                          educ == 13 ~ 1,
                          educ > 13 ~ 2), 
  educ_dummy = factor(educ_dummy, 
                       levels = c(0, 1, 2), 
                      labels = c("low", "medium", "high")), 
  medium_educ = ifelse(educ==13, 1, 0), 
  high_educ = ifelse(educ>13, 1, 0),
  low_educ = ifelse(educ<13, 1, 0)
)

# contrasts(nls_80$educ_dummy)


```

When we include all the education dummies, Stata/R drops `low_educ` because of **singularity** with the other educ dummies; one is a linear combination of the other, in fact perfectly correlated: `corr(high_educ, medium_educ, low_educ)` = `r cor(nls_80[, c("high_educ", "medium_educ", "low_educ")])[1, ]`



\begin{equation}
wage = \beta_0 + \beta_1married + \beta_2high\_iq + \beta_4high\_educ + z\gamma + u\\
\label{mod.2}
\end{equation}


```{r, warning=FALSE, results="asis"}

eqn.1 <- substitute(wage ~ married + high_iq + high_educ + medium_educ + low_educ + exper + age + tenure + feduc)

mod.1 <- lm(eqn.1, nls_80)

# drops low_educ because of singularity, as one is a linear combination of the other. See cor(nls_80[, c("high_educ", "medium_educ", "low_educ")])

eqn.2 <- substitute(wage ~ married + high_iq + high_educ + exper + age + tenure + feduc) # we drop medium_educ

mod.2 <- lm(eqn.2, nls_80)




eqn.3 <- substitute(wage ~ married + high_iq + married:high_iq + high_educ + exper + age + tenure + feduc)

mod.3 <- lm(eqn.3, nls_80)


stargazer(mod.1, mod.2, mod.3,
  title = "Regression output: level-level", 
  dep.var.labels = "wage", font.size = "small",
  omit.stat = c("aic", "bic"), header = FALSE,
  no.space = TRUE, align = TRUE
)
```



(e) \color{red}{Why would you like to interact the married-dummy with the high\_iq-dummy? Run the regression including the interaction term and interpret the coefficient.}



(f) \color{red}{Consider now the a standard log(wage)-equation:}


```{r, results="asis"}
eqn.4 <- substitute(lwage ~ married + educ + exper + I(exper**2) + age + tenure + feduc)

eqn.5 <- substitute(lwage ~ married + log(educ) + exper + I(exper**2) + age + tenure + feduc)

eqn.6 <- substitute(wage ~ married + log(educ) + exper + I(exper**2) + age + tenure + feduc)

mod456 <- list(eqn.4, eqn.5, eqn.6)
names(mod456) <- paste0("mod.", 4:6)
mod456.result <- lapply(mod456, function(x) lm(x, data = nls_80))


stargazer(mod456.result, font.size = "small",
  title = "Regression output: log-level, log-log and level-log", 
  dep.var.labels = c( "ln(wage)", "wage"),
  omit.stat = c("aic", "bic"), header = FALSE,
  no.space = TRUE, align = TRUE
)
```




