---
title:  "Labor Supply: Wages and Hours Worked, Ziliak (1997)"
author: "Eyayaw Beze"
date:   "`r format.Date('2020-07-05', '%B %d, %Y')`"
output: pdf_document
---


```{r setup, include =FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>")
if (!requireNamespace("Ecdat")) install.packages("Ecdat")
library(dplyr)

my_round <- function(x, digits = 2) {
  round(x, digits = 2)
}

labor_ss <- Ecdat::LaborSupply
names(labor_ss)[names(labor_ss)=="lnhr"] <- "lnhrs"
labor_ss <- relocate(labor_ss, c("id", "year"), .before = "lnhrs")
```

# Data Summary
Data on **`r knitr::combine_words(names(labor_ss))`**

```{r }
(N <- length(unique(labor_ss$id))) # cross-sections
(TT <- length(unique(labor_ss$year))) # length of Time
from <- min(labor_ss$year)
to <- max(labor_ss$year)
(NT <- N*TT) # total number of observations
mainvars <- c("lnhrs", "lnwg")
K <- 1 # num of dep var; we lnwg as the only regressor
grand_mean <- my_round(sapply(labor_ss[mainvars], mean))
```

The dimension of the data set is N = `r N` cross-sections by T = `r TT` (years) from `r from` to `r to`. Thus, there are $N\times T = `r N*TT`$ observations. 
The sample means of lnhrs and lnwg are respectively, `r grand_mean[[1]]` and `r grand_mean[[2]]`, implying geometric means of `r format(exp(grand_mean[[1]]), big.mark = ",", digits = 2)` hours and $`r my_round(exp(grand_mean[[2]]))`. 

```{r summary-stats}

summ_stats <- sapply(labor_ss[-grep("^id|year", names(labor_ss))], 
                     function(x, na.rm = FALSE) list(min = min(x),
                                                    mean = mean(x), 
                                                    sd = sd(x),
                                                    max = max(x)), na.rm = TRUE)
dims <- dim(summ_stats)
dimnames <- dimnames(summ_stats)
summ_stats <- my_round(unlist(summ_stats))
dim(summ_stats) <- dims
dimnames(summ_stats) <- dimnames

summ_stats
```

The sample standard deviations are respectively, `r summ_stats[[2]]` and `r summ_stats[[4]]` indicating considerably greater variability in percentage terms in wages rather than hours.  

# Decompose total variation of a series 
We can decompose the total variation of a series $x_{it}$ around its grand mean $\bar x$ into **within sum of squares** and **between sum of squares**: 

$$\begin{aligned} \sum_{i=1}^{N} \sum_{t=1}^{T}\left(x_{i t}-\bar{x}\right)^{2} &=\sum_{i=1}^{N} \sum_{t=1}^{T}\left[\left(x_{i t}-\bar{x}_{i}\right)+\left(\bar{x}_{i}-\bar{x}\right)\right]^{2} \\ &=\sum_{i=1}^{N} \sum_{t=1}^{T}\left(x_{i t}-\bar{x}_{i}\right)^{2}+\sum_{i=1}^{N} \sum_{t=1}^{T}\left(\bar{x}_{i}-\bar{x}\right)^{2} \end{aligned}$$. 

This leads to within standard deviation $S_{\mathrm{W}}$ and between standard deviation $s_{\mathrm{B}}$, where 
\[s_{\mathrm{W}}^{2}=\frac{1}{N T-N} \sum_{i=1}^{N} \sum_{t=1}^{T}\left(x_{i t}-\bar{x}_{i}\right)^{2} \] and \[ s_{\mathrm{B}}^{2}=\frac{1}{N-1} \sum_{i=1}^{N}\left(\bar{x}_{i}-\bar{x}\right)^{2} \]. 

```{r estimation}
 df_between <- 
   labor_ss %>% 
   group_by(id) %>% 
   summarise(across(contains("ln"), mean, .names = "mean_{col}")) %>% 
    select(id, contains("mean_"))
 
 df_within <- 
   labor_ss %>% 
   group_by(id) %>% 
   mutate(across(contains("ln"), function(x) x-mean(x), .names = "demeaned_{col}")) %>%
   select(id, year, contains("demeaned_")) %>% 
    ungroup()
 df_fd <- 
    labor_ss %>% 
    group_by(id) %>% 
    mutate(across(contains("ln"), function(x) x-dplyr::lag(x), .names = "fd_{col}")) %>%
    select(id, year, contains("fd_")) %>% 
    ungroup()

ssw <- sapply(df_within[c("demeaned_lnhrs", "demeaned_lnwg")], 
              function(x) sum(x**2)) 
ssb <- sapply(df_between[c("mean_lnhrs", "mean_lnwg")], 
              function(x) sum((x - mean(x))**2))

sst <- sapply(labor_ss[c("lnhrs", "lnwg")], 
              function(x) sum((x - mean(x))**2))
ss <- rbind(ssw, ssb, sst) # sum of squares between, within and total variations.
dimnames(ss)[[2]] <- mainvars # in terms of the original names of variables
ss 
rm(sst, ssb, ssw)


```
```{r helper}
touch_names <- function(nms, left = TRUE) {
   if (left) 
      sub("(\\w+)_(\\w+)", "\\2", nms) # if stat name first e.g mean_x
   else  
      sub("(\\w+)_(\\w+)", "\\1", nms) # if stat name second e.g. x_mean
   }

# to have identical variable names (against POLS) for stargazer output
names(df_between) <- touch_names(names(df_between))
names(df_within) <- touch_names(names(df_within))
names(df_fd) <- touch_names(names(df_fd))

denominator <- c(NT-N, N-1)
sd_lnhrs <- Map(function(x, y) sqrt(x/y), ss[-3, 1], denominator)
sd_lnwg <- Map( function(x, y) sqrt(x/y), ss[-3, 2], denominator)
```

The within and between standard deviations are, respectively, `r sd_lnhrs[[1]]` and `r sd_lnhrs[[2]] ` for lnhrs, and  `r sd_lnwg[[1]]` and `r sd_lnwg[[2]]`.

```{r }
POLS <- lm(lnhrs~lnwg, labor_ss)
Between <- lm(lnhrs~lnwg, df_between)
Within <- lm(lnhrs~lnwg+0
             , df_within)
fd <- lm(lnhrs~lnwg, df_fd) # first difference 
```

RE variance matrix estimate requires consistent estimates of the variance components $\sigma_{\varepsilon}^{2}$ and $\sigma_{\alpha}^{2}$. From the within or fixed effects regression of $\left(y_{i t}-\bar{y}_{i}\right)$ on $\left(\mathbf{x}_{i t}-\overline{\mathbf{x}}_{i}\right)$ we obtain \[\widehat{\sigma}_{\varepsilon}^{2}=\frac{1}{N(T-1)-K} \sum_{i} \sum_{t}\left(\left(y_{i t}-\bar{y}_{i}\right)-\left(\mathbf{x}_{i t}-\overline{\mathbf{x}}_{i}\right)^{\prime} \widehat{\boldsymbol{\beta}}_{\mathrm{W}}\right)^{2}.\]
From the between regression of $\bar{y}_{i}$ on an intercept and $\overline{\mathbf{x}}_{i},$ an equation that has error term with variance $\sigma_{\alpha}^{2}+\sigma_{\varepsilon}^{2} / T,$ we obtain \[ \widehat{\sigma}_{\alpha}^{2}=\frac{1}{N-(K+1)} \sum_{i}\left(\bar{y}_{i}-\widehat{\mu}_{\mathrm{B}} -\overline{\mathbf{x}}_{i}^{\prime} \widehat{\boldsymbol{\beta}}_{\mathrm{B}}\right)^{2}-\frac{1}{T} \widehat{\sigma}_{\varepsilon}^{2} \]

Then, $\lambda=1-\frac{\sigma_{\varepsilon}}{\left(T \sigma_{\alpha}^{2}+\sigma_{\varepsilon}^{2}\right)^{1 / 2}}$.


```{r }
# variance components of the RE GLS estimator
(var_epsilon <- sum((df_within[["lnhrs"]] - 
                        df_within[["lnwg"]]*coef(Within))**2)/(NT-N-K)
)
(var_alpha <- sum((df_between[["lnhrs"]]- coef(Between)[[1]] - 
                      df_between[["lnwg"]]*coef(Between)[[2]])**2)/(N-(K+1)) - 
      (1/TT) * var_epsilon)

(lambda <- 1 - (var_epsilon)**.5/ (TT * var_alpha + var_epsilon)**.5)
```

The individual effects $\alpha_{it}$ can be estimated by 
$$ \hat\alpha_{i} = \bar y_i - \mathbf{x}_i'\mathbf{\beta_W}.$$ 

```{r }
# estimate individual effects in FE or within estimator
alpha <- df_between[["lnhrs"]]-df_between[["lnwg"]]*coef(Within)
alpha_bar <- mean(alpha)
```

Thus, $\hat\alpha_{i} =$ `r my_round(alpha_bar)`. 

```{r }
coeffs <- lapply(list(POLS, Between, Within, fd), coef) # all coeffs comparison
coeffs[[3]] <- c("(Intercept)" = alpha_bar, coef(Within))
coeffs <- do.call(cbind, coeffs)
covariate.labels <- c("$\\beta$", "$\\alpha$")
dimnames(coeffs)[[1]] <- rev(covariate.labels)
dimnames(coeffs)[[2]] <- c("POLS" ,   "Between", "Within" , "First Diff")
knitr::kable(coeffs)


```
```{r regression-output, results = "asis"}
#
# how to add coefficients by myself? 
# coef = list("$\\alpha$" =coeffs[1,], "$\\beta$" = coeffs[2,])

 stargazer::stargazer(list(POLS, Between, Within, fd),
                      covariate.labels = covariate.labels,
                      dep.var.caption = "",
                      dep.var.labels = "",
                      column.labels = c("POLS", "Between", "Within", "First Diff"), 
                      header = FALSE,
                      model.numbers = FALSE,
                      digits = 3L,
                      font.size = "small",
                      df = FALSE,
                      omit.stat = "adj.rsq",
                      title = "Hours and Wages: Standard Linear Panel Model Estimators")
 
```
```{r convert it into .Rmd, include = FALSE}
knitr::spin("Ziliak-1997.R", knit = FALSE)
```

